---
title: "Biking Example"
author: "Kelly Melton"
date: "2023-06-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(caret)
library(readxl)
library(rpart)
library(dplyr)
library(rpart.plot)
library(ggplot2)
library(tidyverse)
library(zoo)
library(lubridate)
library(forecast)
library(faraway)
library(xts)

BikeEx <- read_excel("C:/Users/kmelton/OneDrive - High Technology Foundation/Desktop/Biking/biking_data.xlsx")

summary(BikeEx)
```


```{r}
BikeEx$dteday <- ymd(BikeEx$dteday)

plot(BikeEx$dteday,BikeEx$windspeed)
plot(BikeEx$dteday,BikeEx$temp)
plot(BikeEx$dteday,BikeEx$hum)
plot(BikeEx$dteday,BikeEx$weathersit)



plot(BikeEx$cnt,BikeEx$windspeed)
plot(BikeEx$cnt,BikeEx$temp)
plot(BikeEx$cnt,BikeEx$hum)
plot(BikeEx$cnt,BikeEx$weathersit)
```


```{r}
cordata = BikeEx[,c(9,10,12,13,16)]
corr <- round(cor(cordata), 1)
corr

```

```{r}
library(ggcorrplot)
ggcorrplot(corr, hc.order = TRUE, type = "lower", lab = TRUE, lab_size = 3, method="circle", colors = c("yellow", "orange", "red"), outline.color = "gray", show.legend = TRUE, show.diag = FALSE, title="Correlation")
```


```{r}
data=BikeEx[,c(9,10,12,13,16)]
scaled_bike <- as.data.frame(scale(data))
summary(scaled_bike)
```

```{r}
m1<-glm(cnt~windspeed + temp + hum + weathersit, data=scaled_bike)
summary(m1)
plot(m1)

#When interpreting the coefficients for TMAX and TMIN, it is more appropriate to focus on the incremental change in the response variable (COUNT) associated with a one-unit increase in these temperature variables, while holding other variables constant. For example, a coefficient of 4.7289 for TMAX suggests that, on average, for every one-unit increase in the maximum temperature, the COUNT variable increases by 4.7289 units, assuming all other variables in the model remain the same.

#It's important to keep in mind the limitations of interpreting coefficients in the presence of complex variables like temperature, where zero is not the baseline reference point.
```
```{r}

# Creating a new dataframe
bikedf<- data.frame(scaled_bike)


# Example prediction for new_data_frame

predictions <- predict(m1, newdata = bikedf, type = "response")
summary(predictions)

```
```{r}
set.seed(300)

smplbike <- sample(1:nrow(scale_data), size=round(0.8*nrow(scale_data)), replace=FALSE)

BikeExtrain <- scaled_bike[samp,]
BikeExtest <- scaled_bike[-samp,]

dim(BikeExtrain)
dim(BikeExtest)
summary(BikeExtrain)


```
```{r}
m2<-glm(cnt ~ windspeed + temp + hum + weathersit, data=BikeExtrain)
summary(m2)
plot(m2)
```


```{r}
predict <- predict(m2, BikeExtrain, type = 'response')
table_mat <- table(BikeExtrain$cnt, predict > 0.5)
table_mat
```


```{r}
accuracy_Test <- sum(diag(table_mat)) / sum(table_mat)
accuracy_Test
```

```{r}
library(xts)
library(tsbox)

#xts will create a time series object
historical = xts(BikeWV[,c("temp","hum","windspeed","weathersit")],order.by=as.Date(BikeEx$dteday))

#na.fill ensures there are no missing dates which will be needed during decomposition 
historical = na.fill(historical, "extend")

#window sets up dates to be aggregated into months later in analysis
historical = window(historical, start=as.Date("2011-01-01"), end=as.Date("2012-12-31"))

```



```{r}

# plot the max and min temp and dates to see if there are dips and jumps, will use an additive model later on if there are dips/jumps


plot(ts_ts(historical$hum), col="pink", bty="n", las=1, fg=NA, 
	ylim=c(-21, 305), ylab="Humidity")

lines(ts_ts(historical$temp), col="blue")

grid(nx=NA, ny=NULL, lty=1, col="gray")

legend("topright", fill=c("pink)
	legend=c("temp", "hum"), bg="white"))
```

```{r}

barplot(historical$PRCP, border=NA, col="blue", ylim=c(0, 700),
	space=0, bty="n", las=1, fg=NA, ylab="Daily Rainfall (10th/mm)")

grid(nx=NA, ny=NULL, lty=1)

```

```{r}

barplot(historical$TMAX, border=NA, col="blue", ylim=c(0, 400),
	space=0, bty="n", las=1, fg=NA, ylab="Max Temp 10ths/C")

grid(nx=NA, ny=NULL, lty=1)

```


```{r}

barplot(historical$TMIN, border=NA, col="blue", ylim=c(0, 250),
	space=0, bty="n", las=1, fg=NA, ylab="MinTemp C")

grid(nx=NA, ny=NULL, lty=1)

```






```{r}

summary(historical)

```




```{r}
# look for dates with historical weather conditions for each attribute

historical[historical$TMIN == min(historical$TMIN)]

historical[historical$TMAX == min(historical$TMAX)]

historical[historical$PRCP == min(historical$PRCP)]

historical[historical$SNOW == min(historical$SNOW)]

```

```{r}
# look for dates with historical weather conditions for each attribute

historical[historical$TMIN == max(historical$TMIN)]

historical[historical$TMAX == max(historical$TMAX)]

historical[historical$PRCP == max(historical$PRCP)]

historical[historical$SNOW == max(historical$SNOW)]

```



```{r}

sum(is.na(historical$TMAX))
class(historical$TMAX)
ts_data <- ts(historical$TMAX, frequency = 365)



#Decomposition is used to breakdown where seasonality exists

decomposition = stl(ts_ts(historical$TMAX), s.window=365, t.window=2190)

plot(decomposition)

summary(decomposition$time.series[,"trend"])

```




```{r}

#Monthly breakdown of historical data for precipitation

months = split(as.numeric(historical$PRCP), historical$MONTH)

sapply(months, summary)

```


```{r}

#Monthly breakdown of historical data for average wind speed 

months = split(as.numeric(historical$AWND), historical$MONTH)

sapply(months, summary)

```



```{r}
#Decomposition is used to breakdown where seasonality exists

decomposition = stl(ts_ts(historical$TMAX), s.window=365, t.window=2190)

plot(decomposition)

summary(decomposition$time.series[,"trend"])

```


For max temp trend line there is an increase of 0.81 degrees C.
(0.81 * 365/6) = 49.27 tenths of degrees C (49.27/10 = 4.92) or an average daily INCREASE in max temp of 4.92 degrees C per day. 


```{r}

decomposition = stl(ts_ts(historical$PRCP), s.window=365, t.window=2190)

plot(decomposition)

summary(decomposition$time.series[,"trend"])

```


For precipitation, there is a daily decrease in the trend line of -30.68. Using the data description for this parameter's measurements, and the formula above, this translates to an average daily DECREASE in daily precipitation of 18.66 mm (0.73 in). 

The data begins in 2017, the year of a historic weather event. The decrease is expected. 


```{r}

decomposition = stl(ts_ts(historical$TMIN), s.window=365, t.window=2190)

plot(decomposition)

summary(decomposition$time.series[,"trend"])

```


Min temp shows average daily DECREASE of 2.85 degrees C.


```{r}

#the mean function is used to smooth randomness and note trends. 

monthly.tmax = period.apply(historical$TMAX, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = mean)

plot(ts_ts(monthly.tmax), col="red", ylim=c(-21, 400), 
	lwd=3, bty="n", las=1, fg=NA, ylab="TMAX (C)")

grid(nx=NA, ny=NULL, lty=1)
```


```{r}

monthly.prcp = period.apply(historical$PRCP, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = mean)

plot(ts_ts(monthly.prcp), col="green", 
	lwd=3, bty="n", las=1, fg=NA, ylab="Monthly Precipitation (10th/mm)")

grid(nx=NA, ny=NULL, lty=1)

```


```{r}

#Show extremes v. average with area plots
#Max = red
#Avg = yellow
#Min = blue

tmax.mean = period.apply(historical$TMAX, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = mean)

tmax.max = period.apply(historical$TMAX, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = max)

tmin.mean = period.apply(historical$TMIN, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = mean)

tmin.min = period.apply(historical$TMIN, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = min)

tmax.area = c(as.numeric(tmax.max), rev(as.numeric(tmax.mean)))
tavg.area = c(as.numeric(tmax.mean), rev(as.numeric(tmin.mean)))
tmin.area = c(as.numeric(tmin.mean), rev(as.numeric(tmin.min)))

indices = c(index(ts_ts(tmax.mean)), rev(index(ts_ts(tmax.mean))))

plot(NA, xlim=range(indices), ylim=c(-20, 100), 
	lwd=3, bty="n", las=1, fg=NA, ylab="Monthly Temperatures (C)")

polygon(indices, tmax.area, border=NA, col="red")
polygon(indices, tavg.area, border=NA, col="yellow")
polygon(indices, tmin.area, border=NA, col="blue")

grid(nx=NA, ny=NULL, lty=1)
```


```{r}
library(forecast)

# additive model using training data to predict future max temps

training.data = period.apply(historical$TMAX, seq(1, nrow(historical) - 1, 30.4305), max)

#h=60 indicates forecasting the next 5 years

model.tmax = hw(ts_ts(training.data), h=60)

```


```{r}
plot(model.tmax, lwd=3, bty="n", las=1, fg=NA)

grid(nx=NA, ny=NULL, lty=1)
```


The blue line is the trend line and the gray areas represent possible values based on the training data. 


```{r}

# three year forecast for max temp

model.tmax = hw(ts_ts(training.data), h=36)

plot(model.tmax, lwd=3, bty="n", las=1, fg=NA)

grid(nx=NA, ny=NULL, lty=1)
```


```{r}
#precipitation forecast for next five years according to the additive model

training.data = period.apply(historical$PRCP, seq(1, nrow(historical) - 1, 30.4305), max)

model.tprcp = hw(ts_ts(training.data), h=36)

```


```{r}

plot(model.tprcp, lwd=3, bty="n", las=1, fg=NA)

grid(nx=NA, ny=NULL, lty=1)

```


The forecasted precipitation is more difficult to predict with this model presumably due to the extreme coastal weather the state experiences.


```{r}

#ARIMA models use a deeper analysis of training data
#these models take time to run

training.data = ts_ts(historical$TMAX)

parameters = auto.arima(training.data)

parameters
```


```{r}

# three year max temp forecast measured in 10ths of degrees C

arima.model = arima(training.data, order = c(5,0,1), seasonal = list(order=c(0,1,0), period=365))

arima.tmax = forecast(arima.model, 2190)

plot(arima.tmax, lwd=3, bty="n", las=1, fg=NA, 
	xlim=c(2017, 2026), ylab="Monthly High Temperature (C)")

grid(nx=NA, ny=NULL, lty=1)
```


```{r}
historical$WT04 = (historical$TMAX + historical$TMIN) / 2

historical$HDD = ifelse(historical$WT04 < 65, 65 - historical$WT04, 0)

historical$CDD = ifelse(historical$WT04 > 65, historical$WT04 - 65, 0)
```


```{r}
monthly.cdd = period.apply(historical$CDD, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = sum)

monthly.hdd = period.apply(historical$HDD, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = sum)

barplot(merge.xts(monthly.cdd, monthly.hdd), 
	col=c("blue", "yellow"), las=1, fg="white",
	space=0, border=NA)

grid(nx=NA, ny=NULL, lty=1)

legend(x="topright", fill=c("blue", "yellow"), legend=c("CDD", "HDD"), bg="white")
```


```{r}
decomposition = stl(ts_ts(historical$CDD), s.window=365, t.window=7001)

plot(decomposition)
```


```{r}
decomposition = stl(ts_ts(historical$HDD), s.window=365, t.window=7001)

plot(decomposition)
```


```{r}
historical$TAVG = (historical$TMAX + historical$TMIN) / 2

monthly.tavg = period.apply(historical$TAVG, INDEX = seq(1, nrow(historical) - 1, 30.4305), FUN = mean)

estimate.hdd = ifelse(monthly.tavg < 65, 65 - monthly.tavg, 0) * 30

estimate.cdd = ifelse(monthly.tavg > 65, monthly.tavg - 65, 0) * 30

barplot(merge.xts(estimate.cdd, estimate.hdd), 
	col=c("blue", "yellow"), las=1, fg="white",
	space=0, border=NA)

grid(nx=NA, ny=NULL, lty=1)

legend(x="topright", fill=c("blue", "yellow"), legend=c("CDD", "HDD"), bg="white")
```


```{r}

historical$PRCP = ifelse(historical$PRCP > 1, 1, 0)

rain = period.apply(historical$PRCP, INDEX = seq(1, nrow(historical) - 1, 365.25), FUN = sum)

plot(ts_ts(rain), col="blue", 
	lwd=3, bty="n", las=1, fg=NA, 
	ylab="Annual Days with High Precipitation")

grid(nx=NA, ny=NULL, lty=1)

summary(as.numeric(rain))

```


```{r}
decomposition = stl(ts_ts(historical$TMIN), s.window=365, t.window=7001)

plot(decomposition)
```


``` {r}
saveRDS(historical, "2017-2022-historical.rds")
```
